---
title: "ggplot2 Exercises (HVTN workshop 2022)"
author: "<your name>"
output: html_document
date: "2022-10-26"
editor_options: 
  chunk_output_type: console
---

The presentation is here: http://bryanmayer.com/hvtn-ggplot2-workshop/hvtn-ggplot2-workshop.html

# 0 - Setup

```{r setup}
# check getwd() for correct directory, setwd() here if not already here.
# OR: Session -> Set Working Directory -> To Source File Location

library(ggplot2)
mock_bama_example = read.csv("mock_bama_example.csv")

```

# 1 - ggplot2 Background

## Initial examples

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

```

```{r base-r-example}

plot(mpg[mpg$class == "compact", ]$displ, mpg[mpg$class == "compact", ]$hwy, 
     col = "orange", xlim = c(1.5, 7), ylab = "hwy", xlab = "displ")   

points(mpg[mpg$class == "2seater", ]$displ, mpg[mpg$class == "2seater", ]$hwy, col = "red")

legend(x = 5, y = 40, c("compact", "2seater"), col = c("orange", "red"), pch = 1)

```

## Geom examples

```{r}
base = ggplot(data = mpg, aes(x = displ, y = hwy)) 
base + geom_point() + ggtitle("geom_point")
base + geom_line() + ggtitle("geom_line")
base + geom_smooth(se = F) + ggtitle("geom_smooth")
base + geom_count(se = F) + ggtitle("geom_count")
ggplot(data = mpg, aes(x = hwy)) + geom_histogram() + ggtitle("geom_histogram")
ggplot(data = mpg, aes(x = class, y = hwy)) + geom_boxplot() + ggtitle("geom_boxplot")

```

## Layering geometries

```{r}
base_boxplot = ggplot(data = mpg, aes(x = class, y = hwy)) 
base_point = ggplot(data = mpg, aes(x = displ, y = hwy)) 

# se = F removes CI bands for the example
base_point + geom_point() + geom_smooth(se = F) + ggtitle("base_point + geom_point() + geom_smooth()")

base_boxplot + geom_boxplot() + geom_point() + ggtitle("base_boxplot + geom_boxplot() + geom_point()")

```

## aes mappings vs. settings

```{r}

ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point(aes(color = "blue"))

ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) + geom_point()

ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point(color = "blue")

# global settings doesn't affect anything
ggplot(data = mpg, aes(x = displ, y = hwy), colour = "blue") + geom_point()

```

## Continuous vs. discrete (boxplot)

```{r}
ggplot(data = mpg, aes(x = cyl, y = hwy)) + geom_boxplot()
ggplot(data = mpg, aes(x = factor(cyl), y = hwy)) + geom_boxplot()
```

## Exercise 1 - Boxplots with mpg

```{r}

# IN CLASS EXERCISE: REPLACE ... TO GENERATE PLOT

mpg_base = ggplot(...) + ...

mpg_base

```

## Scale examples

### Discrete and manual

```{r}

ggplot(data = mpg, aes(x = displ, y = hwy, colour = factor(cyl))) +
  geom_point() +
  scale_color_manual("cylinder #", breaks = c(4, 5, 6, 8),
                     labels = paste(c(4, 5, 6, 8), "cylinder"),
                     values = c("red", "green", "purple", "black"))

```


```{r scales-named-vectors}

# explicit naming for named vector
# use ` ` for non-standard names
cyl_colors = c(`4` = "red", `5` = "green",`6` = "purple", `8` = "black")

# assign names after making the vector
cyl_labels =  paste(c(4, 5, 6, 8), "cylinder")
names(cyl_labels) = c(4, 5, 6, 8)

ggplot(data = mpg, 
       aes(x = displ, y = hwy, colour = factor(cyl))) +
  geom_point() +
  scale_color_manual("cylinder #", values = cyl_colors,
                     labels = cyl_labels)


```

```{r scales-data-frames}
# alternative approach not discussed in class

cyl_mapping = data.frame(
  breaks = c(4, 5, 6, 8),
  labels = paste(c(4, 5, 6, 8), "cylinder"),
  colors = c("red", "green", "purple", "black")
)

ggplot(data = mpg, 
       aes(x = displ, y = hwy, colour = factor(cyl))) +
  geom_point() +
  scale_color_manual("cylinder #", values = cyl_mapping$colors,
                     breaks = cyl_mapping$breaks, labels =  cyl_mapping$labels)

```

### Continuous scales - log plots

```{r logpl-vs-pllog}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  scale_y_log10(breaks = c(15, 20, 30, 40))

ggplot(data = mpg, aes(x = displ, y = log10(hwy))) +
  geom_point() +
  scale_y_continuous(breaks = log10(c(15, 20, 30, 40)))

```

## Coordinate functions (zooming)

```{r}

set.seed(10)
coord_ex = data.frame(grp = rep(LETTERS[1:2], each = 10), y = rnorm(20))

ggplot(coord_ex, aes(x = grp, y = y))+ geom_boxplot() + scale_y_continuous(limits = c(0, NA))
ggplot(coord_ex, aes(x = grp, y = y))+ geom_boxplot() + coord_cartesian(ylim = c(0, NA))
```

## Exercise 2 - Improving exercise 1

```{r}

# IN CLASS EXERCISE: REPLACE ... TO GENERATE PLOT
# USE mpg_base CREATED IN EXERCISE 1

mpg_final = mpg_base +
  ...
  theme_classic()

mpg_final

```

## Faceting Exercise

```{r}

## INTERACTIVE EXERCISE

mpg_final + 
  facet_wrap(...)

```

# 2. HVTN Magnitude Figure

## Explore data and setup

```{r}

dplyr::glimpse(mock_bama_example)

hvtn_pl_colors = c(T1 =  "#1749FF", T2 = "#D92321", T3 = "#0AB7C9", P = "#787873", Negative = "#8F8F8F")
hvtn_pl_shapes = c(T1 =  16, T2 = 16, T3 = 16, P = 16, Negative = 2)
axis_grp_order = c("T1", "T2", "T3", "P")

```

## Exercise 3 - HVTN Magnitude plot

```{r}

# IN CLASS EXERCISE: REPLACE ... TO GENERATE PLOT

ggplot(mock_bama_example, aes(...)) +
  geom_boxplot(..., show.legend = F, outlier.colour = NA) +
  geom_point(..., position = position_jitter(width = 0.2)) +
  scale_shape_manual(values = c(2, 16)) +
  scale_color_manual(values = hvtn_pl_colors) +
  scale_x_discrete(limits = ...) +
  scale_y_...() +
  facet_wrap(...) +  
  labs(...) +
  theme_classic() 


```

## EXERCISE 4 - New variable for color

```{r}

mock_bama_example$grp_response = ifelse(mock_bama_example$response == 0, 
                                         "Negative", 
                                         mock_bama_example$rx_code)


```

```{r}

## COPY AND PASTE ANSWER FROM EXERCISE 3, 
## USE grp_response for color mapping

```

## Finalize 

```{r}

mag_final = ggplot(data = mock_bama_example, 
                   aes(x = rx_code, y = pmax(100, delta), color = grp_response)) +
  stat_boxplot(data = subset(mock_bama_example, response == 1), 
               geom = "errorbar", width = 0.5, show.legend = F) +
  geom_boxplot(data = subset(mock_bama_example, response == 1), 
               show.legend = F, outlier.colour = NA) +
  geom_point(aes(shape = grp_response, group = rx_code), 
             position = position_jitter(width = 0.2)) +
  scale_colour_manual("", values = hvtn_pl_colors) +
  scale_shape_manual("", values = hvtn_pl_shapes) +
  scale_x_discrete(limits = axis_grp_order) +
  scale_y_log10(breaks = c(100, 1000, 2000, 3000, 5000, 10000),
                label = c("< 100", 1000, 2000, 3000, 5000, 10000)) +
  facet_wrap(~ visitno) +
  labs(y = "IgG binding antibody units (MFI - blank)", x = "") +
  theme_classic() +
  theme(strip.background = element_blank(), panel.background = element_rect(color="black"),
        legend.position = "bottom", axis.ticks.x = element_blank(), strip.text = element_blank())

mag_final


```

# 4. Remaining Components of HVTN Figure

## Responder tallies

### Summarize data first

```{r}
library(dplyr)
response_rate_summary = mock_bama_example %>%
  group_by(rx_code, antigen, visitno) %>%
  summarize(
    Total = n(),
    Responders = sum(response),
    rr = mean(response),
    .groups = "drop"
  )

```

### Make "plot"

The trick is involves plotting using geom_text then stripping almost every plot element.

```{r}

rr_txt_pl_skeleton = ggplot(response_rate_summary, 
                            aes(x = rx_code, y = 1, 
                                label = paste(Responders, Total, sep = "\n"))) +
  geom_text(size = 3) +
  scale_x_discrete(limits = axis_grp_order) +
  scale_y_continuous(breaks = 1, labels = "Responders\nTotal", expand = c(0, 0)) +
  facet_wrap(~visitno) +
  labs(y = "") 

rr_txt_pl_skeleton

rr_txt_pl = rr_txt_pl_skeleton +
  theme(panel.background = element_blank(), axis.text.x = element_blank(),
        axis.title.x = element_blank(), axis.ticks = element_blank(),
        strip.background = element_blank(), strip.text = element_blank(), 
        axis.text.y = element_text(size = 10), plot.margin = margin())


rr_txt_pl
```

## Response rate plot

For both approaches. Use of the `labeller` to get the correct names for the facet labels. Alternatively, the visitno variable can be used to create a new variable in the data that uses these levels.

### Approach 1

Uses the stat_summary_bin function. Using a trick that response proportion = mean(response). Equivalent: mean(100*response) = response %.

```{r}

resp_rate1 = ggplot(mock_bama_example, aes(x = rx_code, y = 100*response, fill = rx_code)) +
  stat_summary_bin(fun = "mean", geom = "bar") +
  scale_fill_manual(values = hvtn_pl_colors, guide = "none") +
  scale_x_discrete(limits = axis_grp_order) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 102)) +
  labs(y = "Response rate (%)", x = "") +
  facet_wrap(~visitno, labeller = 
               as_labeller(c(`5` = "Visit 5",
                             `7` = "Visit 7",
                             `11` = "Visit 11"))) +
  theme_classic() +
  theme(strip.background = element_blank(), panel.background = element_rect(color="black"),
        plot.title = element_text(size = 12)) + 
  ggtitle("HVTN Trial IgG Antibody Responses to ag_b")


resp_rate1

```

### Approach 2

Use summarized data.

```{r resp_rate_alt}

resp_rate2 = ggplot(response_rate_summary, aes(x = rx_code, y = 100*rr, fill = rx_code)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = hvtn_pl_colors, guide = "none") +
  scale_x_discrete(limits = axis_grp_order) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 102)) +
  facet_wrap(~visitno, labeller = 
               as_labeller(c(`5` = "Visit 5",
                             `7` = "Visit 7",
                             `11` = "Visit 11"))) +
  labs(y = "Response rate (%)", x = "") +
  theme_classic() +
  theme(strip.background = element_blank(), panel.background = element_rect(color="black"))

resp_rate2

```


## Putting it together

```{r}
library(cowplot)

# need to turn off x-axis in bar plot.
plot_grid(
  resp_rate1 + theme(axis.text.x = element_blank(),  
                     axis.ticks.x = element_blank()), 
  rr_txt_pl +theme(plot.margin = margin(t = -20)),  #handles annoying clipping issue
  mag_final, 
  rel_heights = c(18, 1.75, 30), 
  nrow = 3, align = "v", axis = "lr")


# to get the SOURCE text appended on the bottom:
 # + draw_figure_label(label = paste("SOURCE: SCHARP bmayer ", Sys.time()), 
 #    size = 12, position = "bottom.left")

```



